Wormtable: A data structure for genome scale data

### 2013-07-04 docs/paper/paper.tex
Added some timings

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples of wormtable
that illustrate the improvements over existing methods. We applied these 
examples to a large, publicly available VCF file produced as part of the
"Drosophila Genetic Reference Panel". This VCF ("freeze2.vcf.gz, available 
here: ftp://ftp.hgsc.bcm.edu/DGRP/freeze2_Feb_2013/) contains variant calls 
from a population consisting 192 inbred Drosophila lines and contains data 
from the whole genome over a total of 6,146,611 rows). This VCF was converted to 
wormtable format using the included \texttt{vcf2wt} conversion program.  
The (uncompressed) VCF file was 15Gb, and the converted wormtable was 12Gb,
which required X hours.

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows whereas methods that use the VCF directly 
must parse each row before it can be used, and is by far the most time 
consuming part of processing VCF data. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable and using PyVCF[citation?], a module for
python to parse VCF files. In both of these examples we proceed 
row-by-row, checking the REF and ALT columns for the type of variant and
summing transitions and transversions (ignoring indels). The 
PyVCF method required X hours to complete, while the wormtable 
based script needed only Y seconds. This stark difference is not a 
criticism of the PyVCF library, which provides an efficient parser 
for VCF files; it is rather a measure of the inherent slowness of
parsing text in large volumes.

@@

To illustrate the use of indexes in wormtable, we created an index 
on the REF and ALT columns. This index required X MB of disc space 
Y minutes to build, respectively. The REF+ALT index provides a much more 
efficient means of counting the number of transitions and transversions 
(we can just count the number of rows matching keys in the index 
corresponding the SNPs of each type). A script to calculate these 
counts by this method required x seconds.

>>>>

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples of wormtable
that illustrate the improvements over existing methods. We applied these 
examples to a large, publicly available VCF file produced as part of the
"Drosophila Genetic Reference Panel". This VCF ("freeze2.vcf.gz, available 
here: ftp://ftp.hgsc.bcm.edu/DGRP/freeze2_Feb_2013/) contains variant calls 
from a population consisting 192 inbred Drosophila lines and contains data 
from the whole genome over a total of 6,146,611 rows). This VCF was converted to 
wormtable format using the included \texttt{vcf2wt} conversion program.  
The (uncompressed) VCF file was 15GB, and the converted wormtable was 12GB 
and took 30 minutes, 55 seconds to create.

Values are stored in wormtable in a binary format, so that no parsing is
required when reading in rows whereas methods that use the VCF directly 
must parse each row before it can be used, and is by far the most time 
consuming part of processing VCF data. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset, using wormtable and using PyVCF[http://pyvcf.readthedocs.org/?], a module for
python to parse VCF files. In both of these examples we proceed 
row-by-row, checking the REF and ALT columns for the type of variant and
summing transitions and transversions (ignoring indels). The 
PyVCF method required X hours to complete, while the wormtable 
based script needed only Y seconds. This stark difference is not a 
criticism of the PyVCF library, which provides an efficient parser 
for VCF files; it is rather a measure of the inherent slowness of
parsing text in large volumes.

@@

To illustrate the use of indexes in wormtable, we created an index 
on the REF and ALT columns. This index required 66MB of disc space 
3 minutes 50 seconds to build, respectively. The REF+ALT index provides a much more 
efficient means of counting the number of transitions and transversions 
(we can just count the number of rows matching keys in the index 
corresponding the SNPs of each type). A script to calculate these 
counts by this method required x seconds.
